
---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---



This is a summary of my analysis of the fitness activity data.

The goal was to analyse the total number of steps taken per day and per time interval, over a 2 month period (October and November 2012)

## Loading and preprocessing the data
For the first part of the analysis, I unzipped the raw data and read the csv-formatted data. As the first part of the analysis will ignore NA values, I seperate the data set into those elements that contain NA values and those that do not.

```{r}
fname <- "activity.zip"
if (!file.exists(fname))
    print("Cannot find zip data file in local directory")
unzip(fname)
## After unzipping, read the data in CSV format
activity <- read.csv("activity.csv")
values <- activity[!is.na(activity$steps),]
na_values <- activity[is.na(activity$steps),]
```

## What is the average daily activity pattern?

For the initial analysis of the data, NA data is ignored. I use the non-NA data to plot a histogram that shows the total number of steps taken per day. 

```{r, echo}
## Calculate the total number of steps by date
daily_totals <- data.frame(unlist(lapply(split(values, values$date), function(x) sum(x$steps))))
names(daily_totals) <- "total_steps"
hist(as.numeric(daily_totals$total_steps), breaks = 20, xlim = c(0,25000), ylim = c(0,12), col = "red",
    main = "Histogram of number of steps in a day \n (Oct/Nov 2012)",
	  xlab = "Number of steps in a day")

```

The histogram shows that there were 10 days when less than 1,000 steps were taken, and only 2 days when more than 20,000 steps were taken.

As a next step, I calculate and report the mean and median values
```{r}
mn <- mean(daily_totals$total_steps)
print(paste("Mean number of daily steps is", mn))
md <- median(daily_totals$total_steps)
print(paste("Median number of daily steps is", md))
```
The mean number of steps taken in a day is approximately `r as.integer(mn)`, while the median of the number of daily steps is `r md`, as also shown in the updated histogram below.
```{r}
hist(as.numeric(daily_totals$total_steps), breaks = 20, xlim = c(0,25000), ylim = c(0,12), col = "red",
  main = "Histogram of number of steps in a day \n (Oct/Nov 2012)",
  xlab = "Number of steps in a day")
abline(v = mean(daily_totals$total_steps), col = "blue", lwd = 2)
abline(v = median(daily_totals$total_steps), col = "green", lwd = 2)
legend("topright", legend = c("mean","median"), 
       col = c("blue", "green"), lwd = 2)
```

## Calculating the average daily activity pattern
Next, I analysed the mean number of steps taken per 5-minute time interval. To correctly display the values, the interval values must be tranformed from integers into time values.
```{r}
## Calculate the mean number of steps by time interval
interval_means <- data.frame(unlist(lapply(split(values, values$interval), function(x) mean(x$steps))))
names(interval_means) <- "mean_steps"
intervals <- strptime(sprintf("%04d", as.numeric(unique(values$interval))),        format="%H%M")
plot(intervals, as.numeric(interval_means$mean_steps), type = "l",
     xlab = "Time Intervals", ylab = "Number of steps per interval",
     main = "Mean number of steps per 5-minute time intervals")
```

I now calculate the time interval which has the highest mean number of steps.
```{r}
## Calculate the time interval with the highest number of steps
## Split the data set by time interval, and calculate the total number
## of steps per time interval
interval_totals <- data.frame(unlist(lapply(split(values, values$interval), function(x) sum(x$steps))))
names(interval_totals) <- "total_steps"
## Find the row with the highest number of steps
x <- which.max(interval_totals$total_steps)
y <- max(interval_totals$total_steps)
## Retrieve the name of that row, and display it as a time
n <- rownames(interval_totals)
max <- strptime(sprintf("%04d", as.numeric(n[x])), format="%H%M")
t <- as.POSIXlt(max)
print(paste("Time interval with highest total number of steps is ",t$hour,":", t$min, sep=""))
```

The time interval with the highest total number of steps occurs at `r t$hour`:`r t$min`, and a total of `r y` steps were taken during that time interval over the observed period.

## Imputing missing values

In the analysis so far, I have ignored time intervals that contained NA values. I calculate the number of NA values in the original data set.
```{r}
## How many time intervals have NA values ?
s <- nrow(na_values)
print(paste("There are",s,"time intervals with NA values"))
```

I create a new data set in which the NA value for a given time interval is replaced by the mean value for the same interval on other days. To achieve this, I replace the NA values in the na_values data set by the mean value for the equivalent time interval, and then combine the resulting data set with the original non-NA values.
```{r}
## Replace each NA value by the mean value for the same interval on other dates
for (k in 1:nrow(na_values)) {	
		t <- na_values[k,"interval"]
		na_values[k,"steps"] <- interval_means[as.character(t),"mean_steps"]
		}
## Create a new data set thaat contains the replaced NA values
activity2 <- rbind(values, na_values)
```

We can now draw a new histogram of total daily steps, and calculate the updated mean and median number of daily steps, using the updated data set, to see the impact of replacing NA values with calculated values.
```{r}
## Calculate the total number of steps by date
daily_totals2 <- data.frame(unlist(lapply(split(activity2, activity2$date), function(x) sum(x$steps))))
names(daily_totals2) <- "total_steps"
mn2 <- mean(daily_totals2$total_steps)
print(paste("Mean number of daily steps is", mn2))
md2 <- median(daily_totals2$total_steps)
print(paste("Median number of daily steps is", md2))
hist(as.numeric(daily_totals2$total_steps), breaks = 20, xlim = c(0,25000), ylim = c(0,20), col = "red",
    main = "Histogram of number of steps in a day \n(replacing NA values)",
    xlab = "Number of steps in a day")
```

We can see that, having replaced the NA values with calculated values, the mean number of steps taken in a day has increased to approximately `r as.integer(mn2)`, and that the median of the number of daily steps is identical to the mean (`r as.integer(md2)`). Also, there are now only 2 days with fewer than 1000 steps.

## Are there differences in activity patterns between weekdays and weekends?
For the final part of the analysis, I split the data set into weekdays (Monday to Friday) and weekends (Saturday and Sunday), to see if average daily activity pattern different between weekdays and weekends.
```{r}
## Define a function to determine the day of the week
wd <- function(x) if(weekdays(as.Date(x)) %in% c("Saturday", "Sunday")) 
  "WEEKEND" else "WEEKDAY"

## Split the updated data set by weekday and weekend
activity2$wd <- lapply(activity2$date, wd)
options(stringsAsFactors = TRUE)
wd_activity <- activity2[(activity2$wd == "WEEKDAY"),]
we_activity <- activity2[(activity2$wd == "WEEKEND"),]
```

Now re-draw the mean number of steps per time interval for those two subsets, to compare the daily activity pattern for weekends and weekdays
```{r}
wd_means <- data.frame(unlist(lapply(split(wd_activity, wd_activity$interval), function(x) mean(x$steps))))
names(wd_means) <- "mean_steps"
we_means <- data.frame(unlist(lapply(split(we_activity, we_activity$interval), function(x) mean(x$steps))))
names(we_means) <- "mean_steps"
par(mfrow=c(2,1), mar = c(3,3,2,1), oma = c(2,2,2,0))

plot(intervals, as.numeric(we_means$mean_steps), type = "l", col = "blue",
  main = "Weekend", cex.main = 0.8,
	ylab = "", xlab = "", 
	ylim = c(0,250), cex.axis = 0.75)

plot(intervals, as.numeric(wd_means$mean_steps), type = "l", col = "blue",
	main = "Weekday", cex.main = 0.8,
	ylab = "", xlab = "", 
	ylim = c(0,250), cex.axis = 0.75)

mtext("Comparison of weekend and weekday activity patterns", side = 3, outer = TRUE)
mtext("Number of steps", side = 2, outer= TRUE)
mtext("Time intervals", side = 1, outer = TRUE)

```

As we can see, the daily activity pattern has shifted to the right (i.e. to later in the day) on weekend days, compared to weekdays.